\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-10-31 metatheory-condensed.tex
figure

W-types~\cite{martin1984intuitionistic} are a succinct way to model
inductive types in MLTT;
together with the identity type $\TyId{\cdot}{\cdot}$, they can express a
whole host of inductive types~\cite{hugunin2021not}.
%
A constructor is given by a pair of types
$\goodType{\Gm}{A}{}$ and $\goodType{\Gm,A}{B}{}$ (rule \ruleref{wsig/add}):
given two arguments $\goodTerm{\Gm}{t_1}{A}$ and
$\goodTerm{\Gm,\sub{B}{\SubstExt{\SubstWeak{0}}{t_1}}}{t_2}{T}$,
$\Wsup{\wsig}{t_1}{t_2}$ constructs a term of type~$T$ (rule \ruleref{tm/wsup}).
As the rules suggest, our formulation additionally supports
\emph{W-type signatures}.
A well formed W-type signature \goodWSig{\Gm}{\wsig}{n} is
composed of $n$ pairs of types, each of which represents a constructor
(rules \ruleref{wsig/empty} and \ruleref{wsig/add}),
with $\wsigproj{i}{1}{\wsig}$ and $\wsigproj{i}{2}{\wsig}$ giving the
$i$-th pair in~$\wsig$, and $\wcode{\wsig}$ giving the code of the W-type (\ruleref{tm/w}).

@@

In contrast, $\wsig$ should be hidden from the typing context of any
term that does not invoke $\Wsup{\wsig}{\cdot}{\cdot}$ or $\Wrec{\wsig}{\cdot}{\cdot}$, so that
the term can be reused---without being rechecked---for a different
W-type signature~$\wsig'$ that extends~$\wsig$ with additional
constructors.
Moreover, the typing of the term should be made parametric to
the definitions of those fields that invoke $\texttt{Wsup}$ or $\texttt{Wrec}$,
so that the term can be reused---without being rechecked---when those fields
are overridden to support the extended signature~$\wsig'$.
%
Such abstraction required by family polymorphism is supported via linkages.

@@

This type~$A$ abstracting the context in \ruleref{l/add} and
\ruleref{lsig/add} is not necessarily the same as those recorded by~$\lsig$,
because a field representing the code $\wcode{\wsig}$ of a W-type has to
expose different types to different succeeding fields.
Succeeding fields that invoke
$\Wsup{\wsig}{\cdot}{\cdot}$ or $\Wrec{\wsig}{\cdot}{\cdot}$
should see the concrete signature~$\wsig$, as rules \ruleref{tm/wsup}
and \ruleref{tm/wrec} stipulate, whereas $\wsig$ should be hidden
from all other fields so that they can be reused
for a W-type signature~$\wsig'$ extending~$\wsig$.

We illustrate this with \cref{fig:stlc-linkage-typing}.
On the left is code excerpted from \cref{fig:stlc-mechanized}, and on the right
is how the corresponding fields are modeled and typed in \TT.
%
$\lkg_i$ is the linkage that adds the $i$-th field~$t_i$ with the typing
$\goodTerm{A_i}{t_i}{T_i}$.
%
Here, we feel free to use a notation with explicit names.
We name the context as $\mathit{self}$ (cf.~\cref{fig:stlc-compiled})
and feel free to write its type, which is a dependent tuple type, as a
dependent record type with explicit labels.
%
\begin{itemize}[labelsep=*,leftmargin=1pc]

@@

\item
Similarly, $\wsig_{\texttt{tm}}$ is exposed to the typing context of
$t_{10}$, which models the recursive function \lsti{subst} by invoking
the recursor $\Wrec{\wsig_{\texttt{tm}}}{\cdot}{\cdot}$.
%
Other fields automatically generated by the plugin, such as partial
recursors and various propositional equalities, can also be modeled in
this way by invoking $\texttt{Wrec}$.

@@

Inheritance and code reuse can already be expressed through projection
of fields out of linkages and their inclusion into new linkages.
To make common patterns of linkage manipulations more convenient,
\TT builds \emph{linkage transformers}, whose well-formedness judgments
have form $\goodInh{\Gm}{h}{\lsig_1}{\lsig_2}$.
Linkage transformers are second-class; the only way to use them is to
apply them to linkages to form new linkages (\ruleref{tm/inh}).
%
Derived families can be modeled as linkage transformers.
%
Linkage transformers are inductively constructed from several introduction
rules.
\cref{fig:typing-selected} shows the beta rule of an exemplary transformer,
\ruleref{lxeq/ov/beta}.
It can read as follows: given a linkage of form $\LkgAdd{\lkg}{t_1}$,
applying the transformer $\InhOv{h}{t_2}$ overrides the linkage's last
field~$t_1$ with~$t_2$.
$\InhOv{\InhId}{\wcode{\wsig'_{\texttt{tm}}}}$, for instance, can be used
in \cref{fig:stlc-linkage-typing} to form a derived family that
overrides $\wsig_{\texttt{tm}}$ with an extended
signature~$\wsig'_{\texttt{tm}}$.

@@

\begin{figure}
\small
\newcolumntype{L}{>{$}l<{$}}
\newcolumntype{C}{>{$}c<{$}}
\newcolumntype{R}{>{$}r<{$}}
\begin{tabular}{@{}lR@{\ }C@{\ }LR@{\ \,}C@{\,\ }L@{}}
\lsti!Family STLC.!
& \lkg_0 & \coloneqq & \LkgEmp
\\
\lsti!FInductive tm :=!
& \lkg_1 & \coloneqq & \LkgAdd{\lkg_0}{\dadada}
& \mathit{self} : []
& \vdash
& \wcode{\wsig_{\texttt{tm}}} : \TyS{\wcode{\wsig_{\texttt{tm}}}}
\\
\lsti!| tm_unit : tm | !\dadada
& \lkg_2 & \coloneqq & \LkgAdd{\lkg_1}{\dadada}
& \mathit{self} : [\texttt{tm}:\TyS{\wcode{\wsig_{\texttt{tm}}}}]
& \vdash
& \Wsup{\wsig_{\texttt{tm}}}{\top}{\bot} : \El{\mathit{self.}\texttt{tm}}
\\
\lsti!FRecursion subst !\dadada
\\
\lsti!Case tm_unit := !\dadada
& & &
& \mathit{self} : [\texttt{tm}:\cU, \texttt{tm\_unit}: \El{\texttt{tm}}, \dadada]
& \vdash
& \dadada : \CaseSig{\top}{\bot}{\texttt{id}\to\El{\textit{self.}\texttt{tm}}\to\El{\textit{self.}\texttt{tm}}}
\\
\lsti!End subst.!
& & &
& \mathit{self} : [\texttt{tm}:\TyS{\wcode{\wsig_{\texttt{tm}}}}, \texttt{tm\_unit}: \El{\texttt{tm}}, \dadada]
\\
\lsti!!
\end{tabular}
\caption{Typing of each field}
\label{fig:stlc-linkage-typing}
\end{figure}

>>>>

W-types~\cite{martin1984intuitionistic} are a succinct way to model
inductive types in MLTT;
together with the identity type $\TyId{\cDot}{\cDot}$, they can express a
whole host of inductive types~\cite{hugunin2021not}.
%
A constructor is given by a pair of types
$\goodType{\Gm}{A}{}$ and $\goodType{\Gm,A}{B}{}$ (rule \ruleref{wsig/add}):
given two arguments $\goodTerm{\Gm}{t_1}{A}$ and
$\goodTerm{\Gm,\sub{B}{\SubstExt{\SubstWeak{0}}{t_1}}}{t_2}{T}$,
$\Wsup{\wsig}{t_1}{t_2}$ constructs a term of type~$T$ (rule \ruleref{tm/wsup}).
As the rules suggest, our formulation additionally supports
\emph{W-type signatures}.
A well formed W-type signature \goodWSig{\Gm}{\wsig}{n} is
composed of $n$ pairs of types, each of which represents a constructor
(rules \ruleref{wsig/empty} and \ruleref{wsig/add}),
with $\wsigproj{i}{1}{\wsig}$ and $\wsigproj{i}{2}{\wsig}$ giving the
$i$-th pair in~$\wsig$, and $\wcode{\wsig}$ giving the code of the W-type (\ruleref{tm/w}).

@@

In contrast, $\wsig$ should be hidden from the typing context of any
term that does not invoke $\Wsup{\wsig}{\cDot}{\cDot}$ or $\Wrec{\wsig}{\cDot}{\cDot}$, so that
the term can be reused---without being rechecked---for a different
W-type signature~$\wsig'$ that extends~$\wsig$ with additional
constructors.
Moreover, the typing of the term should be made parametric to
the definitions of those fields that invoke $\texttt{Wsup}$ or $\texttt{Wrec}$,
so that the term can be reused---without being rechecked---when those fields
are overridden to support the extended signature~$\wsig'$.
%
Such abstraction required by family polymorphism is supported via linkages.

@@

This type~$A$ abstracting the context in \ruleref{l/add} and
\ruleref{lsig/add} is not necessarily the same as those recorded by~$\lsig$,
because a field representing the code $\wcode{\wsig}$ of a W-type has to
expose different types to different succeeding fields.
Succeeding fields that invoke
$\Wsup{\wsig}{\cDot}{\cDot}$ or $\Wrec{\wsig}{\cDot}{\cDot}$
should see the concrete signature~$\wsig$, as rules \ruleref{tm/wsup}
and \ruleref{tm/wrec} stipulate, whereas $\wsig$ should be hidden
from all other fields so that they can be reused
for a W-type signature~$\wsig'$ extending~$\wsig$.

We illustrate this with \cref{fig:stlc-linkage-typing}.
On the left is code excerpted from \cref{fig:stlc-mechanized}, and on the right
is how the corresponding fields are modeled and typed in \TT.
%
$\lkg_i$ is the linkage that adds the $i$-th field~$t_i$ with the typing
$\goodTerm{A_i}{t_i}{T_i}$.
%
For readability, we feel free to use a notation with explicit names
in this typing judgment:
we name the context as $\mathit{self}$ (cf.~\cref{fig:stlc-compiled})
and feel free to write its type, which is a dependent tuple type, as a
dependent record type with labels.
%
\begin{itemize}[labelsep=*,leftmargin=1pc]

@@

\item
Similarly, $\wsig_{\texttt{tm}}$ is exposed to the typing context of
$t_{10}$, which models the recursive function \lsti{subst} by invoking
the recursor $\Wrec{\wsig_{\texttt{tm}}}{\cDot}{\cDot}$.
%
Other fields automatically generated by the plugin, such as partial
recursors and various propositional equalities, can also be modeled in
this way by invoking $\texttt{Wrec}$.

@@

Inheritance and code reuse can already be expressed through projection
of fields out of linkages and their inclusion into new linkages.
To make common patterns of linkage manipulations more convenient,
\TT builds \emph{linkage transformers}, whose well-formedness judgments
have form $\goodInh{\Gm}{h}{\lsig_1}{\lsig_2}$.
Linkage transformers are second-class; the only way to use them is to
apply them to linkages to form new linkages (\ruleref{tm/inh}).
%
Derived families can be modeled as linkage transformers.
%
Linkage transformers are inductively constructed from several introduction
rules.
\cref{fig:typing-selected} shows the beta rule of an exemplary transformer,
\ruleref{lxeq/ov/beta}.
It can be read as follows: given a linkage of form $\LkgAdd{\lkg}{t_1}$,
applying the transformer $\InhOv{h}{t_2}$ overrides the linkage's last
field~$t_1$ with~$t_2$.
$\InhOv{\InhId}{\wcode{\wsig'_{\texttt{tm}}}}$, for instance, can be used
in \cref{fig:stlc-linkage-typing} to form a derived family that
overrides $\wsig_{\texttt{tm}}$ with an extended
signature~$\wsig'_{\texttt{tm}}$.

@@

\input{stlc-linkage-typing.tex}
